---
title: "Estimating CATE using HAL"
format: html
---

In this example, we demonstrate one way to estimate the conditional average treatment effect (CATE) function using HAL.

## Problem setup

We consider the following problem setup:    

Suppose our observed data contains $n$ i.i.d copies of the random variable $O=(W,A,Y)\sim P_0\in\mathcal{M}$, where $W\in\mathbb{R}^d$ is a vector of baseline covariates, $A\in\{0,1\}$ is a binary treatment, and $Y\in\mathbb{R}$ is the outcome of interest. The statistical model $\mathcal{M}$ is nonparametric.

Our goal is to estimate the CATE function $\tau_0=E_0(Y\mid A=1,W)-E_0(Y\mid A=0,W)$.

In this example, we use the R-loss $L_{m_0,g_0}(\tau)$ indexed by two nuisance parameters $m$ and $g$ for learning CATE:
$$
L_{m,g}(\tau)=\left( Y-m(W)-(A-g(1\mid W))\tau\right)^2,
$$
where
$$
m(W)=E(Y\mid W)\quad\text{and}\quad g(1\mid W)=P(A=1\mid W).
$$

This loss function is double robust in the sense that we have $\tau_0=\arg\min_{\tau}P_0 L_{m,g}(\tau)$ if either $m=m_0$ or $g=g_0$. 

### Loading required packages

We first load the required packages.
```{r, message=FALSE}
library(hal9001)
library(glmnet)
library(plotly) # for interactive plots
```

### Simulating data
We simulate data from the following data generating mechanism:
- $W=(W_1,W_2,W_3)\sim \text{Uniform}(-1,1)$;     
- $A\sim \text{Bernoulli}(\text{expit}(-0.25W_1+W_2))$;     
- $Y\sim N(\mu(A,W),1)$, where $\mu(A,W)=-0.5+W_1+0.5W_2+0.3W_3+A[0.5W_1+\sin(2\pi W_2)]$.
```{r}
source("dgps/sim_data_cate.R")
set.seed(1234)
n <- 2000
data <- sim_data_cate(n)
W <- data[, grep("W", colnames(data))]
A <- data$A
Y <- data$Y
```

### Use HAL to estimate nuisance functions
First, we use HAL to estimate the nuisance function $m(W)=E(Y\mid W)$:
```{r, cache=TRUE}
# enumerate basis list
basis_list <- enumerate_basis(x = as.matrix(W),
                              max_degree = 2,
                              smoothness_orders = 0)

# make HAL design matrix
phi_W <- make_design_matrix(X = as.matrix(W), blist = basis_list)

# fit HAL
m_fit <- cv.glmnet(x = phi_W, 
                   y = Y,
                   family = "gaussian", 
                   alpha = 1)
m <- as.numeric(predict(m_fit, newx = phi_W, s = "lambda.min", type = "response"))
```

Next, we estimate the other nuisance function $g(1\mid W)=P(A=1\mid W)$:
```{r, cache=TRUE}
# fit HAL
g_fit <- cv.glmnet(x = phi_W, 
                   y = A,
                   family = "binomial", 
                   alpha = 1)
g <- as.numeric(predict(g_fit, newx = phi_W, s = "lambda.min", type = "response"))
```

### Estimate the CATE under the R-loss
```{r, cache=TRUE}
pseudo_outcome <- (Y-m)/(A-g)
pseudo_weights <- (A-g)^2
tau_fit <- cv.glmnet(x = phi_W,
                     y = pseudo_outcome,
                     weights = pseudo_weights,
                     family = "gaussian", 
                     nfolds = 5,
                     alpha = 1)
tau <- as.numeric(predict(tau_fit, newx = phi_W, s = "lambda.min", type = "response"))
```

### Plot the estimated CATE function
Now, let's plot a snapshot of the estimated CATE function as a function of $W_1$ and $W_2$, integrating out $W_3$, and compare it with the true CATE function.
```{r, cache=TRUE}
# CATE on W1, W2
basis_list_W1W2 <- enumerate_basis(x = data[, c("W1", "W2")],
                                   max_degree = 2,
                                   smoothness_orders = 0)
phi_W1W2 <- make_design_matrix(X = as.matrix(data[, c("W1", "W2")]),
                               blist = basis_list_W1W2)
cate_W1W2_fit <- cv.glmnet(x = phi_W1W2, 
                           y = tau,
                           family = "gaussian", 
                           alpha = 1)

# define a grid of W1 and W2 values
W1_vals <- seq(-1, 1, 0.1)
W2_vals <- seq(-1, 1, 0.1)
grid <- expand.grid(W1 = W1_vals,
                    W2 = W2_vals)

# predict CATE on the grid and obtain true CATE on the grid
grid_hal_design <- make_design_matrix(X = as.matrix(grid),
                                      blist = basis_list_W1W2)
grid$pred_cate <- as.numeric(predict(cate_W1W2_fit, newx = grid_hal_design, 
                                     s = "lambda.min", type = "response"))
pred_cate_mat <- matrix(grid$pred_cate, 
                        nrow = length(W1_vals), 
                        ncol = length(W2_vals),
                        byrow = TRUE)
grid$true_cate <- sim_data_cate(n, grid = grid[, c("W1", "W2")])
true_cate_mat <- matrix(grid$true_cate, 
                        nrow = length(W1_vals), 
                        ncol = length(W2_vals),
                        byrow = TRUE)
```

::: {layout-ncol=2}
```{r}
# plot the estimated CATE surface
plt_pred <- plot_ly(x = W1_vals, y = W2_vals, z = pred_cate_mat, 
                    colorscale = "Greens") %>%
  add_surface() %>% 
  layout(title = list(text = "Estimated CATE", 
                      font = list(size = 18, color = "black"),
                      yref = "container", 
                      y = 0.98),
         scene = list(xaxis = list(title = "W1", titlefont = list(size = 14)),
                      yaxis = list(title = "W2", titlefont = list(size = 14)),
                      zaxis = list(title = "CATE", titlefont = list(size = 14), 
                                   range = c(-1.5, 1.5)),
                      camera = list(eye = list(x = -1.5, y = 1.5, z = 0.8))),
         legend = list(x = 0.8, y = 1, font = list(size = 12)))
plt_pred
```

```{r}
# plot the true CATE surface
plt_true <- plot_ly(x = W1_vals, y = W2_vals, z = true_cate_mat, 
                    colorscale = "Reds") %>%
  add_surface() %>% 
  layout(title = list(text = "True CATE", 
                      font = list(size = 18, color = "black"),
                      yref = "container", 
                      y = 0.98),
         scene = list(xaxis = list(title = "W1", titlefont = list(size = 14)),
                      yaxis = list(title = "W2", titlefont = list(size = 14)),
                      zaxis = list(title = "CATE", titlefont = list(size = 14), 
                                   range = c(-1.5, 1.5)),
                      camera = list(eye = list(x = -1.5, y = 1.5, z = 0.8))),
         legend = list(x = 0.8, y = 1, font = list(size = 12)))
plt_true
```
