---
title: "Estimating conditional densities using HAL"
format: html
---

In this example, we demonstrate one possible way to estimate conditional densities using HAL through the pooled logistic regression.

## Problem setup

We observe $n$ i.i.d. copies of $O=(X,T)\sim P_0\in\mathcal{M}$, where

-   $X \in \mathbb{R}^d$ are covariates;

-   $T \in \mathbb{R}$ is a continuous variable.

Our target is the conditional density $p_0(t \mid X)$.

## Pooled logistic regression for estimating conditional densities

One way to estimate the conditional density of $T$, given $X$ using HAL is through discretizing $T$. This technique has been considered in [Diáz and van der Laan (2011)](https://www.degruyterbrill.com/document/doi/10.2202/1557-4679.1356/html). In practice, this can be implemented via a pooled logistic regression. Specifically, consider a set of $k$ bins $\{[t_{j-1},t_j):j=1,\dots,k\}$ formed by $t_0,\dots,t_k$ that partitions $T$. Now, we parameterize the conditional density in terms of discrete hazards for each bin. Let $\tilde{\lambda}_j(X)=P(T\in[t_{j-1},t_j)\mid T\geq t_{j-1},X)$, that is, $\tilde{\lambda}_j$ is the discrete hazard of falling into bin $j$, given having "survived" up to $t_{j-1}$. Then, we have that: $$
P(T\in[t_{j-1},t_j)\mid X)=\tilde{\lambda}_j(X)\prod_{m=1}^{j-1}(1-\tilde{\lambda}_m(X)).
$$ 
The conditional density can then be obtained as
$$
p(t\mid X)=\sum_{j=1}^k \frac{P(T\in[t_{j-1},t_j)\mid X)}{t_j-t_{j-1}}I(t\in[t_{j-1},t_j)).
$$
For $n$ independent units, the likelihood corresponds to the likelihood of an indicator in a repeated measures data set, where the outcome is an indicator of whether $T_i$ falls within a specific bin, i.e., $T_i\in(t_{\alpha-1},t_\alpha]$. For each subject, observations are repeated as many times as there are bins $[t_{\alpha-1},t_\alpha)$ prior to the bin in which $T_i$ lies in. One could then run a logistic regression with, for example, zero-order HAL basis, on this repeated data set to obtain the estimated bin-level conditional densities. Finally, a piecewise constant conditional density function over $\mathbb{R}$ can be obtained by normalizing each bin-level estimate by the corresponding bin width.

Once we have mapped the discrete conditional density into a conditional density, we can use cross-validation based on the log-likelihood loss to select the number of bins $k$.

## Load required packages

We first load the required packages.

```{r, message=FALSE}
library(haldensify)
library(plotly) # for interactive plots
```

The \texttt{R} package \texttt{haldensify} implements the nonparametric conditional density estimation via pooled logistic regressions using HAL, as discussed above \citep{hejazi_efficient_2022,hejazi_haldensify_pkg_2022,hejazi_haldensify_joss_2022}.

## Data generating process

Let's consider a setting where we are tasked to estimate the generalized propensity score (conditional density of a continuous treatment given covariates). This task could arise in the context of, for example, estimating the effect of a stochastic intervention on a continuous treatment. Suppose researchers collect two baseline covariates $W_1$ and $W_2$, each following a standard normal distribution. Suppose the treatment $A$ is a continuous variable following a normal distribution with a mean given by $1+0.3W_1-1.5W_2$. To introduce some heteroskedasticity, we set its standard deviation to $2+0.3|W1|$.

```{r, cache=TRUE}
set.seed(75681)
data <- read.csv("dgps/data_cde.csv")
```

### Use the haldensify function to estimate the conditional density of A given W

The below one liner uses the \texttt{haldensify} function to estimate the conditional density of $A$ given $W=(W_1,W_2)$ using zero-order HAL basis functions. The function automatically selects the number of bins via cross-validation based on the log-likelihood loss.

```{r, cache=TRUE, warning=FALSE}
haldensify_fit <- haldensify(A = data$A,
                             W = data$W,
                             lambda_seq = exp(seq(-0.1, -10, length = 300)))
```

Let's plot the empirical risk (negative log-likelihood) as a function of the regularization parameter $\lambda$.

```{r, cache=TRUE}
# empirical risk vs. lambda
risk_vs_lambda <- plot(haldensify_fit)
risk_vs_lambda
```

## Plot the estimated conditional density

Now, let's visualize the estimated conditional density of $A$ at $W_2=0$ over a grid of $W_1$ values, and compare it with the true conditional density.

```{r}
# compute true conditional densities over a grid
source("dgps/sim_data_cde.R")
n <- 2000
W1_vals <- seq(-2, 2, length.out = 20)
W2_fixed <- 0
A_vals <- seq(-2, 6, length.out = 30)
grid <- expand.grid(W1 = W1_vals,
                    W2 = W2_fixed, 
                    A = A_vals)
grid$true_dens <- sim_data_cde(n, grid = grid)$A
grid$pred_dens <- predict(haldensify_fit, 
                          new_A = grid$A,
                          new_W = grid[, c("W1", "W2")])
pred_dens_mat <- matrix(grid$pred_dens, 
                        nrow = length(A_vals), 
                        ncol = length(W1_vals),
                        byrow = FALSE)
true_dens_mat <- matrix(grid$true_dens, 
                        nrow = length(A_vals), 
                        ncol = length(W1_vals),
                        byrow = FALSE)

plt_pred <- plot_ly(x = A_vals, y = W1_vals, z = pred_dens_mat, 
        colorscale = "Greens") %>%
  add_surface() %>% 
  layout(title = list(text = "Estimated Conditional Density", 
                      font = list(size = 18, color = "black"),
                      yref = "container", 
                      y = 0.98),
         scene = list(xaxis = list(title = "A", titlefont = list(size = 14)),
                      yaxis = list(title = "W1", titlefont = list(size = 14)),
                      zaxis = list(title = "Density", 
                                   titlefont = list(size = 14), 
                                   range = c(0, 0.2)),
                      camera = list(eye = list(x = 1.5, y = 1.5, z = 1))),
         legend = list(x = 0.8, y = 1, font = list(size = 12)))

plt_true <- plot_ly(x = A_vals, y = W1_vals, z = true_dens_mat, 
        colorscale = "Reds") %>%
  add_surface() %>%
  layout(title = list(text = "True Conditional Density", 
                      font = list(size = 18, color = "black"),
                      yref = "container", 
                      y = 0.98),
         scene = list(xaxis = list(title = "A", titlefont = list(size = 14)),
                      yaxis = list(title = "W1", titlefont = list(size = 14)),
                      zaxis = list(title = "Density", 
                                   titlefont = list(size = 14), 
                                   range = c(0, 0.2)),
                      camera = list(eye = list(x = 1.5, y = 1.5, z = 1))),
         legend = list(x = 0.8, y = 1, font = list(size = 12)))
```

::: {layout-ncol="2"}
```{r}
# plot the estimated conditional density
plt_pred
```

```{r}
# plot the true conditional density
plt_true
```
:::
