---
title: "Bulding a diverse super learner library with candidate HAL learners"
format: html
---

### Introduction

### Load required packages

```{r, message=FALSE}
library(hal9001)
library(sl3) # super learner
library(origami) # for V-fold cross-validation
library(tmle)
options(sl3.verbose = TRUE)
```

### Data

```{r}
source("dgps/sim_data_dsl.R")
set.seed(123)
data <- sim_data(500)
W1 <- data$W1
W2 <- data$W2
W3 <- data$W3
A <- data$A
Y <- data$Y
folds <- make_folds(n = nrow(data), V = 5)
```

First, let's build a super learner library for the treatment mechanism. We can use the formula interface in `hal9001` to give HAL specifications.

```{r}
# specify formulas
formula_1 <- ~ h(W1,s=0,k=20) + h(W2,s=0)
formula_2 <- ~ h(W1,s=0,k=20) + h(W2,s=0) + h(W1,W2,s=0,k=10)
formula_3 <- ~ h(W1,s=1,k=20) + h(W2,s=1) + h(W1,W2,s=1,k=10)

# define candidate HAL learners
g_hal_1 <- Lrnr_hal9001$new(formula = formula_1)
g_hal_2 <- Lrnr_hal9001$new(formula = formula_2)
g_hal_3 <- Lrnr_hal9001$new(formula = formula_3)

# define discrete SL library
g_lib <- Stack$new(g_hal_1, g_hal_2, g_hal_3)
g_sl <- Lrnr_sl$new(learners = g_lib, 
                    metalearner = Lrnr_cv_selector$new(loss_loglik_binomial))
```

Next, we define a task for fitting the treatment mechanism.

```{r}
# define task for fitting the treatment mechanism
task_g <- sl3_Task$new(data = data, 
                       covariates = c("W1", "W2", "W3"),
                       outcome = "A",
                       outcome_type = "binomial",
                       folds = folds)

# fit the treatment mechanism and obtain predictions
fit_g <- g_sl$train(task_g)
g1W <- fit_g$predict(task_g)
```

Now, let's build a super learner library for the outcome regression.

```{r}
# specify formulas
formula_Q_1 <- ~ h(., s=0,k=20) + h(.,., s=0,k=20)
formula_Q_2 <- ~ h(W1,s=0,k=20) + h(W2,s=0) + h(W3,s=0,k=20) + 
  h(A,W1,s=0,k=10) + h(A,W3,s=0,k=10)
formula_Q_3 <- ~ h(W1,s=1,k=20) + h(W2,s=1) + h(W3,s=1,k=20) + 
  h(A,W1,s=1,k=10) + h(A,W3,s=1,k=10)

# define candidate HAL learners
Q_hal_1 <- Lrnr_hal9001$new(formula = formula_Q_1)
Q_hal_2 <- Lrnr_hal9001$new(formula = formula_Q_2)
Q_hal_3 <- Lrnr_hal9001$new(formula = formula_Q_3)

# define discrete SL library
Q_lib <- Stack$new(Q_hal_1, Q_hal_2, Q_hal_3)
Q_sl <- Lrnr_sl$new(learners = Q_lib, 
                    metalearner = Lrnr_cv_selector$new(loss_squared_error))
```

Similarly, we define tasks for fitting the outcome regression. Note that for the outcome regression, we need to create three tasks: one for fitting the observed outcome regression $E(Y\mid A,W)$, and two counterfactual tasks for $A=1$ and $A=0$ to obtain predictions under each treatment level.

```{r}
# define task for fitting the outcome regression (and counterfactual tasks)
task_Q <- sl3_Task$new(data = data,
                       covariates = c("A", "W1", "W2", "W3"),
                       outcome = "Y",
                       outcome_type = "continuous",
                       folds = folds)
data_A1 <- data; data_A1$A <- 1; data_A0 <- data; data_A0$A <- 0
task_Q1 <- sl3_Task$new(data = data_A1,
                        covariates = c("A", "W1", "W2", "W3"),
                        outcome = "Y",
                        outcome_type = "continuous",
                        folds = folds)
task_Q0 <- sl3_Task$new(data = data_A0,
                        covariates = c("A", "W1", "W2", "W3"),
                        outcome = "Y",
                        outcome_type = "continuous",
                        folds = folds)

# fit the outcome regression and obtain predictions
fit_Q <- Q_sl$train(task_Q)
QAW <- fit_Q$predict(task_Q)
Q1W <- fit_Q$predict(task_Q1)
Q0W <- fit_Q$predict(task_Q0)
```

Finally, let's supply the nuisance parameter estimate `g1W` and initial estimates `Q1W` and `Q0W` to the `tmle` function to compute the targeted estimates.

```{r}
tmle_obj <- tmle(W = data[, c("W1", "W2", "W3")],
                 Y = data$Y,
                 A = data$A,
                 Q = as.matrix(data.frame(Q0W=Q0W, Q1W=Q1W)),
                 g1W = g1W,
                 family = "gaussian",
                 prescreenW.g	= FALSE,
                 evalATT = FALSE)
summary(tmle_obj)
```

Let's compare the results to the truth:

```{r, cache=TRUE}
truth <- get_truth()
truth
```
